An Empirical Analysis of the Role of Energy in Economic Growth

### 2014-12-30 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Many additional edits.

% The following commands assist making tables for elasticity of substitution (sigma and sigma_1).
% The \firstrow macro creates the sigma_1 row.  The following arguments apply:
% #1 sigma_1 for (kl) nest
% #2 sigma_1 for (kl)e nest
% #3 sigma_1 for (ek)l nest
% #4 sigma_1 for (le)k nest
\newcommand{\firstrow}[4]{\multirow{2}{*}{$\sigma_1$} & ($\underline{kl}$) & #1 
                                  &  & 
                                        & &  \\
                            & ($\underline{kl}$)$e$ & #2
                                  & ($\underline{ek}$)$l$ & #3
                                        & ($\underline{le}$)$k$ & #4 \\
}

% The \secondrow macro creates the sigma row.  The following arguments apply:
% #1 sigma for (kl)e nest
% #2 sigma for (kl)e nest
% #3 sigma for (le)k nest
% #4 sigma for (le)k nest
% #5 sigma for (ek)l nest
% #6 sigma for (ek)l nest
\newcommand{\secondrow}[6]{      \multirow{3}{*}{$\sigma$} & &
                                      & ($\underline{k}l$)$\underline{e}$ & #1
                                            & ($k\underline{l}$)$\underline{e}$ & #2 \\
                                & ($\underline{l}e$)$\underline{k}$ & #3
                                      & ($l\underline{e}$)$\underline{k}$ & #4
                                            &  & \\
                                & ($e\underline{k}$)$\underline{l}$ & #5
                                      & &
                                            & ($\underline{e}k$)$\underline{l}$ & #6 \\
}

@@

@@

@@

@@

**** We need to think about the order of the subsections below. --MKH ****

@@

For the Cobb-Douglas and CES production functions, 
augmentation by energy improves goodness of fit when $\gamma > 0$.%
  \footnote{
  The LINEX function is unhelpful in answering the question 
  ``Is it important to include energy in production functions?'' 
  because LINEX presupposes that energy is the only factor of production.
  }
First, we look at the results of the Cobb-Douglas functions 
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CD_Ternary_Scatter_Graph}. 
Visually, $\gamma > 0$ is shown in the right column 
of Figure~\ref{fig:CD_Ternary_Scatter_Graph} 
as data points that are above the horizontal base of the ternary plots. 
We note that the list of countries for which $\gamma \gg 0$ includes 
Tanzania, South Africa, Saudi Arabia, and the United Kingdom. 
Especially for Tanzania, but also for Saudi Arabia and the United Kingdom, 
it appears certain that including energy improves goodness of fit: 
both the fit to historical data (crosshairs) 
and all resample data points (black dots) 
are above the base of the triangle in the energy columns of Figure~\ref{fig:CD_Ternary_Scatter_Graph}. 
However, for South Africa, significant uncertainty in $\gamma$, 
as shown by the large vertical spread of resample data points, 
leads us to exercise caution in concluding that 
augmenting the Cobb-Douglas production function by energy 
is helpful for improving the goodness of fit to historical data. 
A robust result 
($\gamma$ equal or close to zero and very little vertical spread of the resample points) 
indicates that including energy does not help to explain economic growth in Zambia,
while for United States, Japan, China, and especially Iran, 
it is less certain that augmenting with energy is unhelpful 
(larger vertical spread of resample points).

The results for the CES analysis (Figure~\ref{fig:CES_ternary_graph}) 
indicate that including energy is clearly helpful 
in explaining economic growth in Saudi Arabia, 
because all resample data points are consistently far above the base of the ternary plots 
($\gamma \gg 0$) for all possible ways of nesting of energy with capital and labor. 
Two out of the three possible nestings suggest the importance of including energy for Japan and Tanzania.
For Iran, the crosshairs indicate that energy may explain some growth, 
but this is highly uncertain as the resample data points have a very large vertical spread.  
Explaining economic growth in the United States, China, and Zambia will benefit 
relatively less by including energy, 
but the vertical spreads indicate some uncertainty 
in the cases of the United States and Zambia.  
The results for both the United Kingdom and South Africa 
exhibit significant uncertainty and are contradictory for the various nests.

@@

Some developed economies (United States) benefit little from including energy as a factor of production, 
while some others (United Kingdom) benefit in a minor way. 
For other developed economies (Japan), the results of CD and CES models point in opposite directions.  
For NICs, the results are mixed; 
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa due to lack of precision in determining $\gamma$.  
In one OPEC country, Saudi Arabia, adding energy clearly helps to explain economic growth, 
while in another, Iran, the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (Tanzania) or strong indication otherwise (Zambia).

@@

It has been shown elsewhere~\citep{Ayres:2010ug}
that consideration of energy (specifically, useful work) 
in the production function eliminates the need for the Solow residual~($\lambda$) 
over the very long run (e.g.,~1900--2000). 
The purpose of this paper, however, is to assess common production functions \emph{as used in practice}, 
so we include $\lambda$ terms in all fitted production functions. 
The present approach allows us to answer the question of whether augmentation by 
primary energy in the production function reduces the \emph{value of} the Solow residual,
as opposed to eliminating the \emph{need for} the Solow residual altogether (Ayres and Warr's question).
After augmenting by energy, the Solow-residual ($\lambda$) 
in the Cobb-Douglas~(Figure~\ref{fig:CD_lambda_theta_graph}) 
and CES models~(\ref{fig:CES_lambda_gamma_graph}) 
still explains a part of economic growth in several economies, 
especially China, but also in the United Kingdom, Japan, and Tanzania,  
and to a lesser extent in the United States and Iran. 
In South Africa and Zambia, the Solow-residual is zero, while in Saudi Arabia it is negative.%
  \footnote{
  The negative Solow residual ($\lambda$) for Saudi Arabia is a consequence
  of the fact that all indexed factors of production ($k$, $l$, and $e$) are greater than 
  economic output ($y$) for most of the period under study (1991--2011).
  }
For the Chinese economy, energy augmentation significantly reduces 
the uncertainty of the value of the Solow residual~(Figure~\ref{fig:CES_lambda_gamma_graph}). 
Furthermore, no economies show reduced magnitude of the Solow residual from left to right
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CES_lambda_gamma_graph}.

@@

@@

The choice of CES nesting structure has significant influence 
on model results for both 
relative importance of factors of production ($\delta$ and $\delta_1$ values) and 
substitutability among factors of production ($\sigma$ and $\sigma_1$ values).

@@

For most economies (United Kingdom, Japan, China, South Africa, Saudi Arabia, Iran, and Tanzania), 
Figure~\ref{fig:CES_ternary_graph} shows that the relative importance of 
production factors capital~($k$), labour~($l$), and energy~($e$) 
is different for the three different CES nestings (Equations~\ref{eq:CESkle}--\ref{eq:CESekl}). 
In other words, you cannot view two of the ternary plots 
in the rightmost three columns of Figure~\ref{fig:CES_ternary_graph} and predict the third. 
Nesting structure matters little for the economies of the United States and Zambia.

Certain combinations of economy, parameter values, and CES nesting structure 
come with very large resampling spreads, 
indicating large fitted parameter uncertainties. 
For example, 
for Iran's ($kl$)$e$ nesting in Figures~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph} 
there is virtually no useful information 
in $\delta$, $\delta_1$, $\alpha$, $\beta$, and $\gamma$.

@@

The interpretation of elasticities of substitution 
among factors of production ($k$, $l$, and $e$) is challenging,
because a version of each elasticity of substitution appears three or four times 
among the various nests. 
For example, an elasticity of substitution between capital~($k$) and labor~($l$) 
appears in the following nests: 
($\underline{kl}$), 
($\underline{kl}$)$e$, 
($\underline{l}e$)$\underline{k}$, and 
($e\underline{k}$)$\underline{l}$. 
Tables~\ref{tab:sigma_US}--\ref{tab:sigma_ZM} 
provide interpretation of Figures~\ref{fig:CES_sigma1_delta1_graph}
and~\ref{fig:CES_sigma1_delta1_graph}. 
In the tables,
C (c) indicates strong (weak) complementarity, 
S (s) indicates strong (weak) substitutability, and
$\sim\sim\sim$ ($\sim$) indicates large (small) uncertainty.

@@

For many estimates of elasticity of substitution~($\sigma$ and $\sigma_1$), 
the resampling technique indicates high uncertainty~($\sim\sim\sim$). 
Where results are not uncertain, 
we find complementarity~(C and c) among factors of production 
for most economies with the exception of China 
where an indication of substitutability between labor and energy is observed~(S and s).

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \, \sigma_1 = \infty$), 
is a troubling characteristic of the estimated CES coefficients 
that is rarely reported in the literature.
For some economies (Iran and Zambia), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world 
(United States, United Kingdom, and Japan) exhibits large uncertainty. 
(See United States and United Kingdom rows in Figures~\ref{fig:CD_Ternary_Scatter_Graph}  
and~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph}.)

Some of the uncertainty is explained by the structure of the CES model. 
If $\delta$ or $\delta_1$ achieves an extreme value (0 or 1), 
the corresponding elasticity of substitution ($\sigma$ and $\sigma_1$) becomes meaningless. 
(See Equations~\ref{eq:CES}--\ref{eq:CESekl}.) 
Thus, large uncertainty in $\sigma$ and $\sigma_1$ is expected for any resampled model 
in which parameter estimates are positioned on the left or right boundaries 
of Figures~\ref{fig:CES_sigma1_delta1_graph} or~\ref{fig:CES_sigma_delta_graph}.

@@

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, rigorous statistical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies 
to enhance the reliability of results.

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LinEx) 
and non-linear optimization algorithms (CES), 
the resampling analysis performed herein 
provides insight into the questions of 
(a)~how accurately parameters are known and 
(b)~the implications of parameter imprecision.

**** Move this paragraph elsewhere? --MKH ****
We chose the resampling approach instead of the more-common technique 
of estimating confidence intervals from standard errors for two reasons. 
(1)~Model parameters are highly constrained, and confidence intervals often violate the constraints. 
For example a result of $\alpha = 0.25 \pm 0.3$ is nonsensical, 
because $ \alpha \in [0, 1]$. 
(2)~True confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha = 0.2 + 0.1, -0.05$), 
but the standard error approach yields symmetric confidence intervals. 
The resampling approach that we employed both respects constraints 
and allows for asymmetric confidence intervals.%
  \footnote{
  We have chosen to present parameter uncertainty results visually, 
  both because the resampling technique lends itself to visual presentation and 
  because we believe visual comparison is more effective than tablefuls of numbers.
  }

A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to historical economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2--0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} ~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****

The implications of parameter uncertainties should be obvious, 
but we will state them here. 
Basing policy upon parameters with high uncertainty can lead to bad, or completely wrong, policy. 
**** Martin: can you find an example in our results that drives this point home? --MKH  
May be barking up wrong production factor, may assume subs when complementarity: ****

We caution against using $R^2$ as the only criterion for choosing a model, 
because, in our experience, 
both Cobb-Douglas and CES models yield high $R^2$ values for all countries 
(see Figures~\ref{fig:CD_GDP_Spaghetti_Graph} and~\ref{fig:CES_Spaghetti_Graph}), yet
very different uncertainties for fitted parameters.
In some cases, LINEX models fare poorly regarding $R^2$ and might be excluded on that basis. 
(See SA and TZ in Figure~\ref{fig:LINEX_Spaghetti_Graph}.) 
In our opinion, 
it is equally, if not more, important to know the uncertainty 
of estimated parameters than to assess $R^2$.
With rigorous uncertainty testing as we have done here, 
the quality of parameter estimates is exposed: 
at a glance, the tightness of resampled results indicates 
whether or not confidence in the model and data is warranted. 
We conclude that it is insufficient to argue 
in favor of a particular production function 
on the grounds $R^2$.

@@

Several criteria could be applied 
to the question of which production function is best
for describing the role of energy in economic growth:
best match to historical data, 
minimum ambiguity in model structure,
minimum uncertainty in model parameters, and
easy parameter estimation process.
In the sections below, we evaluate the Cobb-Douglas, CES, and LINEX
models using these criteria.

@@

Figures~\ref{fig:CD_GDP_Spaghetti_Graph}, \ref{fig:CES_Spaghetti_Graph}, and
\ref{fig:LINEX_Spaghetti_Graph} indicate that the CES production function
is consistently better than Cobb-Douglas or LINEX in terms of matching historical data.
Cobb-Douglas is a close second, but
LINEX struggles, especially with the Saudi Arabian and Tanzanian economies.%
  \footnote{
  The order of these results is unsurprising: 
  CES contains 6 fitted parameters; Cobb-Douglas, 4; and LINEX, 3.
  We expect models with more parameters to provide a better fit to historical data.
  }$^,$%
  \footnote{
  We also note that in contrast to our approach, shared by \citet{Warr:2012cg},
  of fitting constant $a_0$ and $c_t$ values, 
  \citet{Kummel:2011aa} used logistic functions to generate time-varying
  $a_0$ and $c_t$ values.
  \citeauthor{Kummel:2011aa}'s approach will necessarily produce a better match to historical data,
  because the logistic curves involve several additional fitting parameters.
  }

@@

The energy-augmented CES production function presents ambiguity to the energy analyst
in the form of the choice of factor nesting.
We showed above that nesting can significantly affect 
the interpretation of macroecnomic dynamics, especially 
the relative importance of factors of production~(\ref{fig:CES_ternary_graph}) and 
the understanding of substitutibility 
among factors of production. (See Section~\ref{sec:elasticity_of_substitution}).
Thus, the CES production function has the undesireable characteristic 
that model structure (i.e., nesting) affects both parameter values 
and macroeconomic interpretation.
Neither the Cobb-Douglas nor the LINEX production function 
exhibits similar structural ambiguity.

Thus, Cobb-Douglas or LINEX are preferred on this criterion.

@@

Uncertainties in fitted parameters give rise to uncertainty 
in macroeconomic dynamics inferred from fitted parameters.
Thus, small uncertainty in fitted parameters is desirable. 
Figures~\ref{fig:CD_lambda_theta_graph}, \ref{fig:CD_Ternary_Scatter_Graph},
\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_ternary_graph}, and
\ref{fig:LINEXq_Resample_Graph} show that 
Cobb-Douglas and LINEX have, by and large, 
smaller uncertainties in fitted parameters than the CES production function.
Thus, Cobb-Douglas and LINEX are preferred on this criterion.

>>>>

% The following commands assist making tables for elasticity of substitution (sigma and sigma_1).
% The \firstrow macro creates the sigma_1 row.  The following arguments apply:
% #1 sigma_1 for (kl) nest
% #2 sigma_1 for (kl)e nest
% #3 sigma_1 for (ek)l nest
% #4 sigma_1 for (le)k nest
\newcommand{\firstrow}[4]{\multirow{2}{*}{$\sigma_1$} & ($\underline{k}$+$\underline{l}$)+() & #1 
                                  &  & 
                                        & &  \\
                            & ($\underline{k}$+$\underline{l}$)+($e$) & #2
                                  & ($\underline{e}$+$\underline{k}$)+($l$) & #3
                                        & ($\underline{l}$+$\underline{e}$)+($k$) & #4 \\
}

% The \secondrow macro creates the sigma row.  The following arguments apply:
% #1 sigma for (kl)e nest
% #2 sigma for (kl)e nest
% #3 sigma for (le)k nest
% #4 sigma for (le)k nest
% #5 sigma for (ek)l nest
% #6 sigma for (ek)l nest
\newcommand{\secondrow}[6]{      \multirow{3}{*}{$\sigma$} & &
                                      & ($\underline{k}$+$l$)+($\underline{e}$) & #1
                                            & ($k$+$\underline{l}$)+($\underline{e}$) & #2 \\
                                & ($\underline{l}$+$e$)+($\underline{k}$) & #3
                                      & ($l$+$\underline{e}$)+($\underline{k}$) & #4
                                            &  & \\
                                & ($e$+$\underline{k}$)+($\underline{l}$) & #5
                                      & &
                                            & ($\underline{e}$+$k$)+($\underline{l}$) & #6 \\
}

@@

We choose the resampling approach instead of the more-common technique 
of estimating confidence intervals from standard errors for two reasons. 
First, model parameters are highly constrained, and confidence intervals often violate the constraints. 
For example a result of $\alpha = 0.25 \pm 0.3$ is nonsensical, 
because $ \alpha \in [0, 1]$. 
Second, true confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha = 0.2 + 0.1, -0.05$), 
but the standard error approach yields symmetric confidence intervals. 
The resampling approach that we employed both respects constraints 
and allows for asymmetric confidence intervals.%
  \footnote{
  In addition, we choose to present parameter uncertainty results visually 
  (see Figures~\ref{fig:CD_lambda_theta_graph}, \ref{fig:CD_Ternary_Scatter_Graph},
  \ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_ternary_graph}, and
  \ref{fig:LINEXq_Resample_Graph}),
  both because the resampling technique lends itself to visual presentation and 
  because we believe visual comparison is more effective than tables of numbers.
  }

@@

<<abg_trajectories, eval=FALSE>>=
# This code can be run to test the conclusion about the LINEX model meeting constraints.
# This chunk of R code has eval=FALSE, because it is not a part of the paper.
Calvin_Models <- loadPostProcessedData(Source = "Calvin", kind = "models", dir=pp_path)
abgplotdata <- fortify(Calvin_Models$Calvin$IR$linex$`iK+iL+iQp`)
triPlot(data=abgplotdata, grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear")
@

@@

**** End other notes ****

@@

**** Martin: Do you think it would be helpful to tie the conclusions 
of each section to the literature? 
If so, give it a shot. --MKH ****

@@

For the Cobb-Douglas and CES production functions, 
augmentation by energy improves goodness of fit when $\gamma > 0$.%
  \footnote{
  The LINEX function is unhelpful in answering the question 
  ``Is it important to include energy in production functions?'' 
  because LINEX presupposes that energy is the only factor of production.
  }
First, we evaluate the results of the Cobb-Douglas functions 
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CD_Ternary_Scatter_Graph}. 
Visually, $\gamma > 0$ is shown in the right column 
of Figure~\ref{fig:CD_Ternary_Scatter_Graph} 
as data points that are above the horizontal base of the ternary plots. 
We note that the list of countries for which $\gamma \gg 0$ includes 
Tanzania, South Africa, Saudi Arabia, and the United Kingdom. 
Especially for Tanzania, but also for Saudi Arabia and the United Kingdom, 
it appears certain that including energy improves goodness of fit: 
both the fit to historical data (crosshairs) 
and all resample data points (black dots) 
are above the base of the triangle in the energy columns of Figure~\ref{fig:CD_Ternary_Scatter_Graph}. 
However, for South Africa, significant uncertainty in $\gamma$, 
as shown by the large vertical spread of resample data points, 
leads us to exercise caution in concluding that 
augmenting the Cobb-Douglas production function by energy 
is helpful for improving the goodness of fit to historical data. 
A robust result 
($\gamma$ equal or close to zero and very little vertical spread of the resample points) 
indicates that including energy does not help to explain economic growth in Zambia,
while for United States, Japan, China, and especially Iran, 
it is less certain that augmenting with energy is unhelpful 
(larger vertical spread of resample points).

The results for the CES analysis (Figure~\ref{fig:CES_ternary_graph}) 
indicate that including energy is clearly helpful 
in explaining economic growth in Saudi Arabia, 
because all resample data points are consistently far above the base of the ternary plots 
($\gamma \gg 0$) for all possible ways of nesting energy with capital and labor. 
Two out of the three possible nestings suggest the importance of including energy for Japan and Tanzania.
For Iran, the crosshairs indicate that energy may explain some growth, 
but this is highly uncertain as the resample data points have a very large vertical spread.  
Explaining economic growth in the United States, China, and Zambia will benefit 
relatively less by including energy, 
but the vertical spreads indicate some uncertainty 
in the cases of the United States and Zambia.  
The results for both the United Kingdom and South Africa 
exhibit significant uncertainty and are contradictory for the various nests.

@@

Some developed economies (United States) benefit little from including energy as a factor of production, 
while some others (United Kingdom) benefit in a minor way. 
For other developed economies (Japan), 
the results of Cobb-Douglas and CES models point in opposite directions.  
For NICs, the results are mixed:
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa due to lack of precision in determining $\gamma$.  
In one OPEC country (Saudi Arabia) adding energy clearly helps to explain economic growth, 
while in another (Iran), the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (Tanzania) or strong indication otherwise (Zambia).

@@

It has been shown elsewhere~\citep{Ayres:2010ug}
that consideration of energy (specifically, useful work) 
in the production function eliminates the need for the Solow residual~($\lambda$) 
over the very long run (e.g.,~1900--2000). 
The purpose of this paper, however, is to assess common production functions \emph{as used in practice}, 
so we include $\lambda$ terms in all fitted production functions. 
The present approach allows us to answer the question of whether augmentation by 
primary energy in the production function reduces the \emph{value of} the Solow residual,
as opposed to eliminating the \emph{need for} the Solow residual altogether 
(\citeauthor{Ayres:2010ug}'s question).
After augmenting by energy, the Solow-residual~($\lambda$) 
in the Cobb-Douglas~(Figure~\ref{fig:CD_lambda_theta_graph}) 
and CES models~(Figure~\ref{fig:CES_lambda_gamma_graph}) 
still explains a part of economic growth for several economies, 
especially China, but also for the United Kingdom, Japan, and Tanzania,  
and to a lesser extent for the United States and Iran. 
In South Africa and Zambia, the Solow-residual is zero, while for Saudi Arabia it is negative.%
  \footnote{
  The negative Solow residual ($\lambda$) for Saudi Arabia is a consequence
  of the fact that all indexed factors of production~($k$, $l$, and $e$) are greater than 
  economic output~($y$) for most of the period under study (1991--2011).
  See Figure~\ref{fig:Historical_Graph}.
  }
For the Chinese economy, energy augmentation significantly reduces 
the uncertainty of the value of the Solow residual~(Figure~\ref{fig:CES_lambda_gamma_graph}). 
Furthermore, no economies show reduced magnitude of the Solow residual from left to right
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CES_lambda_gamma_graph}.

@@

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LINEX) 
and non-linear optimization algorithms (CES), 
the resampling analysis performed herein 
provides insight into the questions of 
(a)~how accurately parameters are known and 
(b)~the implications of parameter imprecision.

A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to historical economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2--0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****

The implications of parameter uncertainties should be obvious, 
but we will state them here. 
Policy based upon parameters with high uncertainty can lead to bad, or completely wrong, policy. 
**** Martin: can you find an example in our results that drives this point home? --MKH  
May be barking up wrong production factor, may assume subs when complementarity: ****

We caution against using $R^2$ as the only criterion for choosing a model, 
because, in our experience, 
both Cobb-Douglas and CES models yield high $R^2$ values for all countries 
(see Figures~\ref{fig:CD_GDP_Spaghetti_Graph} and~\ref{fig:CES_Spaghetti_Graph}), yet
very different uncertainties for fitted parameters.
In some cases, LINEX models fare poorly regarding $R^2$ and might be excluded on that basis. 
(See South Africa and Tanzania in Figure~\ref{fig:LINEX_Spaghetti_Graph}.) 
In our opinion, 
it is equally, if not more, important to know the uncertainty 
of estimated parameters than to assess $R^2$.
With rigorous uncertainty testing as we have done here, 
the quality of parameter estimates is exposed: 
at a glance, the tightness of resampled results indicates 
whether or not confidence in the parameters or model is warranted. 
We conclude that it is insufficient to argue 
in favor of a particular production function 
on the grounds of $R^2$ alone.

@@

The subsections below assess whether the choice of CES nesting structure 
influences model results for both 
relative importance of factors of production ($\delta$ and $\delta_1$ values) and 
substitutability among factors of production ($\sigma$ and $\sigma_1$ values).

@@

For most economies (United Kingdom, Japan, China, South Africa, Saudi Arabia, Iran, and Tanzania), 
Figure~\ref{fig:CES_ternary_graph} shows that the relative importance of 
production factors capital~($k$), labor~($l$), and energy~($e$) 
is different for the three different CES nestings (Equations~\ref{eq:CESkle}--\ref{eq:CESekl}). 
In other words, you cannot view two of the ternary plots 
in the rightmost three columns of Figure~\ref{fig:CES_ternary_graph} and predict the third. 
Nesting structure matters little for the economies of the United States and Zambia.

Certain combinations of economy, parameter values, and CES nesting structure 
come with very large resampling spreads, 
indicating large fitted parameter uncertainties. 
For example, 
for Iran's ($k$+$l$)+($e$) nesting in Figures~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph}, 
there is virtually no useful information 
in $\delta$, $\delta_1$, $\alpha$, $\beta$, and $\gamma$.

Thus, we find that nesting has significant influence on model results.

@@

The interpretation of elasticities of substitution 
among factors of production ($k$, $l$, and $e$) is challenging,
because a version of each elasticity of substitution appears three or four times 
among the various nests. 
For example, an elasticity of substitution between capital~($k$) and labor~($l$) 
appears in the following nests: 
($\underline{k}$+$\underline{l}$), 
($\underline{k}$+$\underline{l}$)+($e$), 
($\underline{l}$+$e$)+$\underline{k}$, and 
($e$+$\underline{k}$)+$\underline{l}$. 
Tables~\ref{tab:sigma_US}--\ref{tab:sigma_ZM} 
provide interpretation of Figures~\ref{fig:CES_sigma1_delta1_graph}
and~\ref{fig:CES_sigma_delta_graph}. 
In the tables,
C (c) indicates strong (weak) complementarity, 
S (s) indicates strong (weak) substitutability, and
$\sim\sim\sim$ ($\sim$) indicates large (small) uncertainty.

@@

For many estimates of elasticity of substitution~($\sigma$ and $\sigma_1$), 
the resampling technique indicates high uncertainty~($\sim\sim\sim$). 
Where results are not uncertain, 
we find complementarity~(C and c) among factors of production 
for most economies with the exception of China, 
where an indication of substitutability between labor and energy is observed.

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \, \sigma_1 = \infty$), 
is a troubling characteristic of the estimated CES coefficients 
that is rarely investigated in the literature.
Indeed, for some economies (Iran and Zambia), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world 
(United States, United Kingdom, and Japan) exhibits large uncertainty.

Some of the uncertainty is explained by the structure of the CES model itself. 
If $\delta$ or $\delta_1$ achieves an extreme value (0 or 1), 
the corresponding elasticity of substitution ($\sigma$ and $\sigma_1$) becomes meaningless. 
(See Equations~\ref{eq:CES}--\ref{eq:CESekl}.) 
Thus, large uncertainty in $\sigma$ and $\sigma_1$ is expected for any resampled model 
in which parameter estimates are positioned on the left or right boundaries 
of Figures~\ref{fig:CES_sigma1_delta1_graph} or~\ref{fig:CES_sigma_delta_graph}.

@@

\vspace{5 mm}

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, rigorous statistical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies 
to enhance the reliability of results. 
**** Would econometrically specifying production function relationships and
performing sensitivity analysis really ``enhance the reliability'' of results? 
Or, would doing so improve transparency? --MKH ****

@@

Several criteria could be applied 
to the question of which production function is best
for describing the role of energy in economic growth. 
We consider four criteria here:
best match to historical data, 
minimum ambiguity in model structure,
minimum uncertainty in model parameters, and
robust parameter estimation process.
In the sections below, we evaluate the Cobb-Douglas, CES, and LINEX
models on these criteria.

@@

Figures~\ref{fig:CD_GDP_Spaghetti_Graph}, \ref{fig:CES_Spaghetti_Graph}, and
\ref{fig:LINEX_Spaghetti_Graph} indicate that the CES production function
is consistently better than Cobb-Douglas or LINEX in terms of matching historical data.
Cobb-Douglas is a close second, but
LINEX struggles, especially with the Saudi Arabian and Tanzanian economies.%
  \footnote{
  The order of these results is unsurprising: 
  energy-augmented CES contains 6 fitted parameters; 
  Cobb-Douglas, 4; 
  and LINEX, 3.
  We expect models with more parameters to provide a better fit to historical data.
  }$^,$%
  \footnote{
  We also note that in contrast to our approach of fitting constant $a_0$ and $c_t$ values, 
  which is shared by \citet{Warr:2012cg},
  \citet{Kummel:2011aa} used logistic functions to generate time-varying
  $a_0$ and $c_t$ values.
  \citeauthor{Kummel:2011aa}'s approach will necessarily produce a better match to historical data,
  because the logistic curves involve several additional fitting parameters.
  }

@@

The energy-augmented CES production function presents 
significant ambiguity to the energy analyst
in the form of the choice of factor nesting.
We showed above that choice of nesting 
can significantly affect 
the interpretation of macroecnomic dynamics, 
especially the relative importance of 
factors of production (see Section~\ref{sec:importance_of_factors}
and Figure~\ref{fig:CES_ternary_graph}) and 
the understanding of substitutibility 
among factors of production (see Section~\ref{sec:elasticity_of_substitution}).
Thus, the CES production function has the undesireable characteristic 
that model structure (i.e., nesting) affects both parameter values 
and macroeconomic interpretation.
Neither the Cobb-Douglas nor the LINEX production function 
presents similar structural ambiguity.

Thus, Cobb-Douglas and LINEX are preferred on this criterion.

@@

Uncertainties in fitted parameters give rise to uncertainty 
in macroeconomic dynamics inferred from fitted parameters.
Thus, small uncertainty in fitted parameters is desirable. 
Figures~\ref{fig:CD_lambda_theta_graph}, \ref{fig:CD_Ternary_Scatter_Graph},
\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_ternary_graph}, and
\ref{fig:LINEXq_Resample_Graph} show that 
Cobb-Douglas and LINEX have, by and large, 
smaller uncertainties in fitted parameters than CES.
Thus, Cobb-Douglas and LINEX are preferred on this criterion.
